---
title: "HMM Functions"
author: "Marina LaForgia"
date: "8/27/2021"
output:
  word_document: default
  pdf_document: default
---

Outstanding questions: 
- If the models are going all the way to the 100th iteration, does that mean it is not converging and we should increase the number of steps?
- What is a decent cutoff for how many plots a species needs to occur in out of 80 sites for a decent estimate? 

# Markov Chain Parameter Functions

## 1. Seedbank transitions
This sets up a 2x2 matrix of seed bank transitions

c = colonization probability
g = germination probability
s = seed survival probability
r = probability of seed production (Pluntz model assumes that if it survives to flowering, it has a 100% chance of producing seed)

* (1,1): 1-c -> probability of going from no seedbank in year t to no seedbank in t+1
* (1,2): c -> probability of going from no seedbank to a seedbank
* (2,1): (1-c)x(1-s)x(1-gr) -> probability of going from a seed bank to no seedbank
* (2,2): 1 - ((1-c)x(1-s)x(1-gr)) -> probability of going from a seed bank to a seedbank

```{r}

# matrix of Y transitions (Y = seedbank)
# Zt (combined state) is a function of the seedbank (Y) at time t-1 and flora (x) at time t)
calculateQY = function(param) # calculate QY is a function of param; QY transition matrix 
{
  result = matrix(0,2,2) # result is a matrix of zeros with 2 columns and 2 rows
  result[1,1] = 1 - param$c # probability of going from A to A: that plants don't colonize
  result[2,1] = (1 - param$c)*(1 - param$s)*(1 - param$g*param$r) # probability of going from P to A: the second row of the first column is (1-c) * (1-s) * (1-gF) = probability that seeds that don't colonize, dont survive, and don't germinate
  result[,2] = 1 - result[,1] # column 2 is the opposite of column one so that each row sums to 1; probability of going from A to P is the probability that seeds colonize
  # probability of going from P to P is 1 - probability that seeds don't colonize, don't survive and don't germinate - we were getting stuck on this before but they all have to sum to one!
  return(result) 
}

```

##  2. State transition probability matrix
Matrix of transitions Zt = (Yt-1, Xt) where Y(t-1) is the state of the seed bank in the previous year, and Xt is the state of standing flora aboveground. the three states in order are (0,0), (1,0), (1,1). where

* Zt = (0,0) = both seeds and standing flora are absent (AA)
* Zt = (1,0) = seeds are present but standing flora is absent the following year (PA)
* Zt = (1,1) = seeds are present and then standing flora is present (PP)

Note: AP does not exist as a state, a seed HAD to have been present in t-1 in order to be present aboveground at time t

```{r cars}
calculateQZ = function(param) # Qz transition matrix
{
  result = matrix(0,3,3) # 3 by 3 matrix of 0s
  result[1,1] = 1 - param$c # AA to AA: probability seeds don't colonize
  result[2,1] = (1 - param$c)*(1 - param$s) # PA to AA: probability seeds don't colonize, and seeds that are there don't survive
  result[3,1] = (1 - param$c)*(1 - param$s)*(1 - param$r) # PP to AA; If r = 1, then this will always be zero. We assume that if a seed germinates and flowers it will produce a seed and if any seed is produced it goes into the seed bank (can't go from PP to AA), so why write out these other params? We can mess with r to see if things change, or we can change this to 0; Interesting because we could have the possibility that a seed is produced, but it colonizes a different location and thus DOESNT enter the seed bank
  # maybe r = one because too many unestimable parameters
  result[,2] = (1 - result[,1])*(1 - param$g) #[1,2] = AA to PA: (1-g)*c; [2,2] = PA to PA = (1-g)(1-(1-c)(1-s)); [3,2] = PP to PA = (1-g)
  result[,3] = (1 - result[,1])*param$g #[1,3] = AA to PP = gc; [2,3] = PA to PP = g(1-(1-c)(1-s)); [3,3] = PP to PP = g
  return(result)
}

```

## 3. Probability invariant of a transition matrix.  

I'm not sure what a probability invariant matrix is or what it's purpose is.
Is this the left eigenvector? the reproductive rate? Either way I think this is calculating the equilibrium state of the system

```{r}

calculatePeq = function(Q)
{
  Vmat = eigen(t(Q))$vectors # extract the eigenvectors from QZ
  #We look for an eigenvector whose components have the same sign and we renormalize it so that the sum makes 1 because all probabilities must sum to 1
  n = dim(Q)[1] # n = number of rows in QZ = number of possible states
  for (i in 1:n)
  {
    v = Vmat[,i] 
    if ( all(sign(v)>=0) | all(sign(v)<=0) ) return(v/sum(v))
  }
}

# finding the equilibirum probability of being in each state, there are 3 states depending on whether we are working with QZ or QY
# stable distribution of states

```


## 4. Putting it all together

Add the object param these additional fields;

Still not sure what PZeq and PYeq, though I suspect the equilibrium state of the system and the seedbank respectively, just not understanding how we get that

```{r}

makeParametersCalculations = function(param)
{
  param$QY = calculateQY(param) # calculate seed bank transitions (QY)
  param$QZ = calculateQZ(param) # calculate state transition probability matrix (QZ)
  param$PZeq = calculatePeq(param$QZ) # calculate the invariant of QZ transition matrix; equilibrium state?
  param$PYeq = param$PZeq[2] + param$PZeq[3] # equilibrium state of seedbank?, this just uses the second and third value of the PZeq vector
  param$pZ0 = c(1-param$p0, param$p0*(1-param$g), param$p0*param$g) # initial distribution vector
  return(param) 
}

```

# Expectation Maximization 

Computes maximum likelihood estimators of g, c, and s

## 1. Forward algorithm for one time series 
Recursive scheme for updating the likelihood and state probabilities of the HMC going forward through time

```{r}

pZXprespast = function(X,param)  # forward algorithm for one time series
{
  tf = length(X) # time frame is length of flora
  result = matrix(0,3,tf) # result is a matrix of 0s with three rows (each state: AA, PA, PP) and tf columns
  for (t in 1:tf) # for each column (time step)
  {
    if (t == 1) pZXpast = param$pZ0 else pZXpast = t(param$QZ) %*% result[,t-1] # if it's the first time step, use the initial distribution vector, otherwise transpose the state transition probability matrix and do matrix multiplication on the result matrix from the previous time step
    result[,t] = pZXpast # paste it in for the current time step
    if (X[t] == 1) result[1:2,t] = 0 else result[3,t] = 0 # if flora observed in the current time step, the result for both rows one and two are 0 (because it's observed so it cant be absent belowground) otherwise the third row is 0)
  }
  return(result)
}
```

## 2. Backward algorithm for one time series
Recursive scheme for updating the likelihood and state probabilities of the HMC going backward in time

```{r}

pXfuturegivenZ = function(X,param) #backward algorithm for one time series
{
  tf = length(X) # number of time steps
  result = matrix(0,3,tf) # same matrix as before
  result[,tf] = 1 # starting at 1
  for (t in tf:2) 
  {
    PZXgivenOldZ = param$QZ # take the state transition probability matrix
    if (X[t] == 1) PZXgivenOldZ[,1:2] = 0 else PZXgivenOldZ[,3] = 0 # if flora is observed in that time step, the first two columns are 0, otherwise the 3rd column is 0
    result[,t-1] = PZXgivenOldZ %*% result[,t]
  }
  return(result)
}
```

## 3. Forward Backward algorithm

Creates a function to calculate the log-likelihood and lambda coefficiencts in vector to feed into the forward backward algorithm

Recursive scheme for calculating state probabilities for any point in time, the first function is for...?

```{r}

vectForwardBackward = function(X,param) #log-likelihood and lambda coefficients for one time series; 
{
  tf = length(X) # length of the time series
  pZXpp = pZXprespast(X,param)
  pXfgZ = pXfuturegivenZ(X,param)
  pZX = pZXprespast(X,param)*pXfuturegivenZ(X,param)   #P(Zt,whole vector X)
  pZZX = array(rep(param$QZ,tf-1),c(3,3,tf-1))           #P(Zt,Zt+1,whole vector X)
  for (j in 1:3) pZZX[,j,] = pZZX[,j,]*pZXpp[,1:(tf-1)]
  for (i in 1:3) pZZX[i,,] = pZZX[i,,]*pXfgZ[,2:tf]
  pZZX[,1:2,X[2:tf] == 1] = 0
  pZZX[,3,X[2:tf] == 0] = 0
  likelihood = mean(apply(pZX,2,sum))
  result = numeric(0)
  result$ll = log(likelihood)
  result$coeffZ = pZX[,1]/likelihood
  result$coeffQ = apply(pZZX,1:2,sum)/likelihood
  return(result)
}

ForwardBackward = function(X,param) #same function extended for a N*T matrix of time series
{
  d = dim(X)
  if (length(d) == 0) return(vectForwardBackward(X,param)) else
  {
    N = d[1]
    result = numeric(0)
    llvect = apply(X,1,function(Xvect) vectForwardBackward(Xvect,param)$ll)  #N-vector,i_th component is the log-likelihood of patch i
    result$ll = sum(llvect)                                                 #sum on the N patches
    coeffZmat = apply(X,1,function(Xvect) vectForwardBackward(Xvect,param)$coeffZ) #3*N-matrix,i-th column is the lambda(z) vector of patch i
    result$coeffZ = apply(coeffZmat,1,sum)    #sum on the N patches
    coeffQarray = array(apply(X,1,function(Xvect) vectForwardBackward(Xvect,param)$coeffQ),c(3,3,N)) #3*3*N-matrix,[,,i] is the lambda(z,z') matrix of patch i
    result$coeffQ = apply(coeffQarray,1:2,sum)       #sum on the N patches
    return(result)
  }
}

```

## 4. Log Likelihood
returns the loglikelihood of the model from the forward backward algorithm

```{r}

logLikelihood = function(X,param) ForwardBackward(X,param)$ll # extract log likelihood from final model

```


## 5. Maximization function

I start to get really lost around here

```{r}
Maximization = function(coeffQ, coeffZ, p0 = 'free', g = 'free', c = 'free', s = 'free', r = 'free')
{
  # takes into account cases of non-identifiability
  if (p0 == 'free') p0 = sum(coeffZ[2:3])/sum(coeffZ)
  if (g == 'free') g = (coeffZ[3] + sum(coeffQ[,3]))/(sum(coeffZ[2:3]) + sum(coeffQ[,2:3]))
  if (r == 'free') c2 = sum(coeffQ[3,2:3])/sum(coeffQ[3,])      #calculation of c double prime (first step)
  if (s == 'free')                                              #calculation of c prime (first step)
  {
    c1 = sum(coeffQ[2,2:3])/sum(coeffQ[2,])
    if (r == 'free') if(c2 < c1) r = 0
    if (r == 0) c1 = sum(coeffQ[2:3,2:3])/sum(coeffQ[2:3,])  
  }
  if (c == 'free')                                              #estimation of c
  {
    c0 = sum(coeffQ[1,2:3])/sum(coeffQ[1,])                     #calculation of c (first step)
    if (s == 'free') if(c1 < c0) s = 0
    if (s == 0)
    {
      if(r == 0) c0 = sum(coeffQ[,2:3])/sum(coeffQ) else c0 = sum(coeffQ[1:2,2:3])/sum(coeffQ[1:2,])
      c1 = c0
    }
    c = c0
  }
  if (s == 'free') if(c1 < c) s = 0 else s = 1 - (1-c1)/(1-c)   #estimation of s
  c1 = 1- (1-c)*(1-s)
  if (r == 'free') if(c2 < c1) r = 0 else r = 1 - (1-c2)/(1-c1) #estimation of r
  result = numeric(0)
  result$p0 = p0
  result$g = g
  result$c = c
  result$s = s
  result$r = r
  return(result)
}

```

## 6. Expected Maximization iteration

```{r}
EMiteration = function(X, param, p0 = 'free', g = 'free', c = 'free', s = 'free', r = 'free')
{
  FBresult = ForwardBackward(X,param)
  coeffZ = FBresult$coeffZ
  coeffQ = FBresult$coeffQ
  newparam = Maximization(coeffQ,coeffZ,p0,g,c,s,r)
  newparam = makeParametersCalculations(newparam)
  EMresult = numeric(0)
  EMresult$ll = FBresult$ll
  EMresult$newparam = newparam
  return(EMresult)
}

```

## 7. Expected maximization estimation

```{r}
EMestimation = function(X, nIterations = 150, precision = 10^(-5), p0 = 'free', g = 'free', c = 'free', s = 'free', r = 'free')
{
  alwaysSeedSurvival = FALSE
  if (s == 1 | c == 1 | (g == 1 & r == 1))
  {
    alwaysSeedSurvival = TRUE
    if (r == 'free') print('r non identifiable')
    r = 1
    if (s == 'free') print('s non identifiable')
    s = 1
  }
  if (p0 == 1 & alwaysSeedSurvival)
  {
    if (c == 'free') print('c non identifiable')
    c = 1
  }
  if (g == 1)
  {
    if (s == 'free' & r == 'free') print('r non identifiable from s')
    r = 0
  }
  lllist = rep(0,nIterations)
  p0list = rep(0,nIterations)
  glist = rep(0,nIterations)
  clist = rep(0,nIterations)
  slist = rep(0,nIterations)
  rlist = rep(0,nIterations)
  param = numeric(0)
  if (p0 == 'free') param$p0 = runif(1) else param$p0 = p0
  if (g == 'free') param$g = runif(1) else param$g = g
  if (c == 'free') param$c = runif(1) else param$c = c
  if (s == 'free') param$s = runif(1) else param$s = s
  if (r == 'free') param$r = runif(1) else param$r = r
  param = makeParametersCalculations(param)
  oldLogLikelihood = -Inf
  for (k in 1:nIterations)
  {
    print(t(param[c('p0','g','c','s','r')]))
    EMresult = EMiteration(X,param,p0,g,c,s,r)
    logLikelihood = EMresult$ll
    lllist[k] = logLikelihood
    p0list[k] = param$p0
    glist[k] = param$g
    clist[k] = param$c
    slist[k] = param$s
    rlist[k] = param$r
    print(logLikelihood)
    if(logLikelihood < oldLogLikelihood + precision) break
    print(k)
    oldLogLikelihood = logLikelihood
    param = EMresult$newparam
  }
  EMresult = numeric(0)
  EMresult$lllist = lllist
  EMresult$p0list = p0list
  EMresult$glist = glist
  EMresult$clist = clist
  EMresult$slist = slist
  EMresult$rlist = rlist
  EMresult$ll = logLikelihood
  EMresult$param = param
  return(EMresult)
  
}


```


# Climate Change (McLaughlin)
Split up the time series to look at whether changes in climate (declining rainfall) in the grassland dataset has contributed to shifts in survival/colonization behavior

2000-2010: higher av rainfall, lower variability
2011-2019: lower av rainfall, higher variability

##1. HMM on 1st half of time series
```{r, echo = F}

mcl.df.beg <- expand.grid(Species_Name = names(species.list), p0 = NA, g = NA, c = NA, s = NA, r = NA) # empty df to fill with rates

for(j in names(species.list)){ # for each species
  X = as.matrix(species.list[[j]][,1:10]) # use their time series PA data first half of time series
  n = 5 # why n = 5?
  p0Results = rep(0,n)
  gResults = rep(0,n)
  cResults = rep(0,n)
  sResults = rep(0,n)
  rResults = rep(0,n)
  llResults = rep(0,n)
  lltrueParam = rep(0,n)
  
  for (i in 1:n) {
    for (k in 1:5) print(i) # I don't understand what's going on here, why print it 5 times?
    print(logLikelihood(X, trueParam)) # print the log-likelihood given the "true params"
    EMresult = EMestimation(X, r = 1) # update the model params, stop when new log likelihood is lower than the old log likelihood plus some precision 
    print(logLikelihood(X, trueParam)) # print the log-likelihood given the "true params"; why is this on here twice?
    # fill in results with new params from the best log likelihood model
    p0Results[i] = EMresult$param$p0 # initial seed bank prob
    gResults[i] = EMresult$param$g # germ
    cResults[i] = EMresult$param$c #col
    sResults[i] = EMresult$param$s #surv
    rResults[i] = EMresult$param$r #prod
    llResults[i] = EMresult$ll # log-likelihood
    lltrueParam[i] = logLikelihood(X,trueParam) # log likelihood of x given "true" parameters
  }

  mcl.df.beg[mcl.df.beg$Species_Name == j,]$p0 <- EMresult$param$p0
  mcl.df.beg[mcl.df.beg$Species_Name == j,]$g <- EMresult$param$g
  mcl.df.beg[mcl.df.beg$Species_Name == j,]$c <- EMresult$param$c
  mcl.df.beg[mcl.df.beg$Species_Name == j,]$s <- EMresult$param$s
  mcl.df.beg[mcl.df.beg$Species_Name == j,]$r <- EMresult$param$r
  
}
```

##2. HMM on 2nd half of time series

```{r, echo = F}
mcl.df.end <- expand.grid(Species_Name = names(species.list), p0 = NA, g = NA, c = NA, s = NA, r = NA) # empty df to fill with rates


for(j in names(species.list)){ # for each species
  X = as.matrix(species.list[[j]][,11:20]) # use their time series PA data
  n = 5 # why n = 5?
  p0Results = rep(0,n)
  gResults = rep(0,n)
  cResults = rep(0,n)
  sResults = rep(0,n)
  rResults = rep(0,n)
  llResults = rep(0,n)
  lltrueParam = rep(0,n)
  
  for (i in 1:n) {
    for (k in 1:5) print(i) # I don't understand what's going on here, why print it 5 times?
    print(j)
    print(logLikelihood(X, trueParam)) # print the log-likelihood given the "true params"
    EMresult = EMestimation(X, r = 1) # update the model params, stop when new log likelihood is lower than the old log likelihood plus some precision 
    print(logLikelihood(X, trueParam)) # print the log-likelihood given the "true params"; why is this on here twice?
    # fill in results with new params from the best log likelihood model
    p0Results[i] = EMresult$param$p0 # initial seed bank prob
    gResults[i] = EMresult$param$g # germ
    cResults[i] = EMresult$param$c #col
    sResults[i] = EMresult$param$s #surv
    rResults[i] = EMresult$param$r #prod
    llResults[i] = EMresult$ll # log-likelihood
    lltrueParam[i] = logLikelihood(X,trueParam) # log likelihood of x given "true" parameters
  }

  # here I assumed that the last estimated parameters, were what the model converged on but I'm not sure that's correct?
  mcl.df.end[mcl.df.end$Species_Name == j,]$p0 <- EMresult$param$p0
  mcl.df.end[mcl.df.end$Species_Name == j,]$g <- EMresult$param$g
  mcl.df.end[mcl.df.end$Species_Name == j,]$c <- EMresult$param$c
  mcl.df.end[mcl.df.end$Species_Name == j,]$s <- EMresult$param$s
  mcl.df.end[mcl.df.end$Species_Name == j,]$r <- EMresult$param$r
  
}

```

##3. Graphical relationships

```{r, echo = F}
mcl.df.beg$tf <- "2000-2010"
mcl.df.end$tf <- "2011-2019"

mcl.df.split <- rbind(mcl.df.beg, mcl.df.end)

ggplot(mcl.df.split, aes(x = s, y = c, group = tf, col = tf)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x)

mcl.df.split <- merge(mcl.df.split, trait[,c(1,3,17,18,19)], by = "Species_Name", all.y = F)
mcl.df.split <- filter(mcl.df.split, Annual.Perennial == "Annual")
mcl.df.split$FunGroup <- paste(mcl.df.split$Native.Exotic, mcl.df.split$Grass.Forb.Shrub)

ggplot(mcl.df.split, aes(x = s, y = c, group = tf, col = tf)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x) +
  facet_wrap(~FunGroup)

ggplot(mcl.df.split, aes(x = s, y = c, group = Species_Name)) +
ggforce::geom_link2(aes(group = Species_Name, color = tf), size = 1, n = 500, lineend = "round") +
  facet_wrap(~FunGroup)

# Colonization seems to decrease but seed survival is not necessarily increasing to compensate... next steps: adding seed traits!

ggplot(mcl.df.split, aes(y = s, x = tf)) +
  geom_boxplot() +
  facet_wrap(~FunGroup)

ggplot(mcl.df.split, aes(y = c, x = tf)) +
  geom_boxplot() +
  facet_wrap(~FunGroup)

```

# Climate Change (Sonoran Desert)
Split up the time series to look at whether changes in climate (declining rainfall) in the sonoran desert dataset has contributed to shifts in survival/colonization behavior

1983-1999: higher av rainfall, lower variability
2000-2016: lower av rainfall, higher variability

##1. HMM on 1st half of time series
```{r, echo = F}

sd.df.beg <- expand.grid(Species_Name = names(species.list), p0 = NA, g = NA, c = NA, s = NA, r = NA) # empty df to fill with rates

for(j in names(species.list)){ # for each species
  X = as.matrix(species.list[[j]][,1:17]) # use their time series PA data from first half
  n = 5 # why n = 5?
  p0Results = rep(0,n)
  gResults = rep(0,n)
  cResults = rep(0,n)
  sResults = rep(0,n)
  rResults = rep(0,n)
  llResults = rep(0,n)
  lltrueParam = rep(0,n)
  
  for (i in 1:n) {
    for (k in 1:5) print(i) 
    print(logLikelihood(X, trueParam)) 
    EMresult = EMestimation(X, r = 1)  
    print(logLikelihood(X, trueParam)) 
    p0Results[i] = EMresult$param$p0 # initial seed bank prob
    gResults[i] = EMresult$param$g # germ
    cResults[i] = EMresult$param$c #col
    sResults[i] = EMresult$param$s #surv
    rResults[i] = EMresult$param$r #prod
    llResults[i] = EMresult$ll # log-likelihood
    lltrueParam[i] = logLikelihood(X,trueParam) # log likelihood of x given "true" parameters
  }

  sd.df.beg[sd.df.beg$Species_Name == j,]$p0 <- EMresult$param$p0
  sd.df.beg[sd.df.beg$Species_Name == j,]$g <- EMresult$param$g
  sd.df.beg[sd.df.beg$Species_Name == j,]$c <- EMresult$param$c
  sd.df.beg[sd.df.beg$Species_Name == j,]$s <- EMresult$param$s
  sd.df.beg[sd.df.beg$Species_Name == j,]$r <- EMresult$param$r
  
}
```

##2. HMM on 2nd half of time series

```{r, echo = F}
sd.df.end <- expand.grid(Species_Name = names(species.list), p0 = NA, g = NA, c = NA, s = NA, r = NA) # empty df to fill with rates


for(j in names(species.list)){ # for each species
  X = as.matrix(species.list[[j]][,18:33]) # use their time series PA data
  n = 5 # why n = 5?
  p0Results = rep(0,n)
  gResults = rep(0,n)
  cResults = rep(0,n)
  sResults = rep(0,n)
  rResults = rep(0,n)
  llResults = rep(0,n)
  lltrueParam = rep(0,n)
  
  for (i in 1:n) {
    for (k in 1:5) print(i) # I don't understand what's going on here, why print it 5 times?
    print(j)
    print(logLikelihood(X, trueParam)) # print the log-likelihood given the "true params"
    EMresult = EMestimation(X, r = 1) # update the model params, stop when new log likelihood is lower than the old log likelihood plus some precision 
    print(logLikelihood(X, trueParam)) # print the log-likelihood given the "true params"; why is this on here twice?
    # fill in results with new params from the best log likelihood model
    p0Results[i] = EMresult$param$p0 # initial seed bank prob
    gResults[i] = EMresult$param$g # germ
    cResults[i] = EMresult$param$c #col
    sResults[i] = EMresult$param$s #surv
    rResults[i] = EMresult$param$r #prod
    llResults[i] = EMresult$ll # log-likelihood
    lltrueParam[i] = logLikelihood(X,trueParam) # log likelihood of x given "true" parameters
  }

  # here I assumed that the last estimated parameters, were what the model converged on but I'm not sure that's correct?
  sd.df.end[sd.df.end$Species_Name == j,]$p0 <- EMresult$param$p0
  sd.df.end[sd.df.end$Species_Name == j,]$g <- EMresult$param$g
  sd.df.end[sd.df.end$Species_Name == j,]$c <- EMresult$param$c
  sd.df.end[sd.df.end$Species_Name == j,]$s <- EMresult$param$s
  sd.df.end[sd.df.end$Species_Name == j,]$r <- EMresult$param$r
  
}

```

##3. Graphical relationships

```{r, echo = F}
sd.df.beg$tf <- "2000-2010"
sd.df.end$tf <- "2011-2019"

sd.df.split <- rbind(sd.df.beg, sd.df.end)

ggplot(sd.df.split, aes(x = s, y = c, group = tf, col = tf)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x)

sd.df.split <- merge(sd.df.split, trait[,c(1,3,17,18,19)], by = "Species_Name", all.y = F)
sd.df.split <- filter(sd.df.split, Annual.Perennial == "Annual")
sd.df.split$FunGroup <- paste(sd.df.split$Native.Exotic, sd.df.split$Grass.Forb.Shrub)

ggplot(sd.df.split, aes(x = s, y = c, group = tf, col = tf)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x) +
  facet_wrap(~FunGroup)

ggplot(sd.df.split, aes(x = s, y = c, group = Species_Name)) +
ggforce::geom_link2(aes(group = Species_Name, color = tf), size = 1, n = 500, lineend = "round") +
  facet_wrap(~FunGroup)

# Colonization seems to decrease but seed survival is not necessarily increasing to compensate... next steps: adding seed traits!

ggplot(sd.df.split, aes(y = s, x = tf)) +
  geom_boxplot() +
  facet_wrap(~FunGroup)

ggplot(sd.df.split, aes(y = c, x = tf)) +
  geom_boxplot() +
  facet_wrap(~FunGroup)

```